---
title: "Get Started with glysmith"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with glysmith}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Glycoverse is an incredibly comprehensive ecosystem.
On one hand, this means you can flexibly perform a wide range of analyses by combining its different components.
On the other hand, getting started can require learning quite a lot.

`glysmith` is the highest-level package in the glycoverse suite.
It encapsulates the most common analysis workflows into a unified interface,
allowing you to execute the entire analytical pipeline in just one function call.

If you've seen The Lord of the Rings, you can think of `glysmith` as the "One Ring to rule them all."

*Note: `glysmith` is not part of the core glycoverse packages. You need to install it separately, even if you have installed the meta package `glycoverse`.*

```{r setup}
library(glysmith)
library(glyexp)
```

## Analyze your data with a single line of code

We will demonstrate using the built-in experiment object from `glyexp`.
This is an N-glycoproteomics dataset containing 12 samples across four different liver conditions:
healthy controls (H), hepatitis (M), cirrhosis (Y), and hepatocellular carcinoma (C),
with 3 samples for each group.
Please note, the dataset is raw and has not been normalized or imputed.

To work with your own data, you can use `glyread` to import outputs from tools like pGlyco3, MSFragger-Glyco, or Byonic;
Alternatively, you can create your own experiment object.
For more details, see [Get Started with glyread](https://glycoverse.github.io/glyread/articles/glyread.html)
or [Creating Experiments](https://glycoverse.github.io/glyexp/articles/create-exp.html).

```{r}
real_experiment
```

Now, let's run the default analysis pipeline with just a single line of code.

```{r}
#| message: false
result <- forge_analysis(real_experiment)
result
```

That's it!
You've completed the following steps in one go:

- Preprocessed your data using `glyclean::auto_clean()` and generated QC plots.
- Summarized the experiment with `glyexp::summarize_experiment()`.
- Performed principal component analysis with `glystats::gly_pca()`.
- Conducted differential expression analysis using `glystats::gly_limma()`.
- Plotted a volcano plot via `glyvis::plot_volcano()`.
- Generated a heatmap for significant glycoforms using `glyvis::plot_heatmap()`.
- Conducted GO enrichment analysis using `glystats::gly_enrich_go()`.
- Performed KEGG enrichment analysis using `glystats::gly_enrich_kegg()`.
- Performed Reactome enrichment analysis using `glystats::gly_enrich_reactome()`.
- Derived site-specific traits using `glydet::derive_traits()`.
- Conducted differential trait analysis with `glystats::gly_limma()`.
- Generated a heatmap of significant site-specific derived traits via `glyvis::plot_heatmap()`.

What a comprehensive analysis pipeline!

## Save all results with one line of code

Once you have your results, what's next?
You can export all tables and plots to a target directory in just a single line of code.

```r
quench_result(result, "path/to/save")
```

Running this will create the folder `path/to/save` with the following content:

- `README.md`: Describes all the saved outputs.
- `experiment.rds`: The processed experiment object.
- `meta.rds`: The metadata list.
- `plots/`: All plots generated from the pipeline.
- `tables/`: All tables output from the analysis.

Give it a try yourself!

## One line to generate a report

We're developing advanced report generation features, but for now,
you can use `polish_report()` to generate a report directly from the results.

```r
polish_report(result, "path/to/save/report.html")
```

By default, the report will open automatically in your browser when it's finished.

## About naming

You may have noticed some unusual names in this package.
First, the 'smith' in `glysmith` doesn't refer to a surname;
it's taken from "blacksmith"—someone who makes and repairs items made of iron.

All functions and classes are named using blacksmith jargon.
For example, `forge_analysis()` for "forging" results,
`quench_result()` for "quenching" the results to save them,
and `polish_report()` for "polishing" the report to improve readability.

You'll learn about "blueprint" in the next section.

## Blueprints

Imagine you visit a blacksmith to order a sword.
If you simply say, "I want a sword," the blacksmith will use a default blueprint and start working on it.
However, you always have the option to bring your own design—a custom blueprint—that the smith can use for your sword.

Similarly, `forge_analysis()` uses a default blueprint for the analysis pipeline.
You can inspect it via `blueprint_default()`.

```{r}
blueprint_default()
```

A blueprint is a list of steps executed sequentially.
You can use `blueprint()` with `step_` functions to create your own custom blueprint.
For instance, if you'd just like to perform differential expression analysis and plot the volcano plot, you could do:

```{r}
#| message: false

bp <- blueprint(
  step_preprocess(),
  step_dea_limma(),
  step_volcano(),
)
result <- forge_analysis(real_experiment, bp)
result
```

You can also save a blueprint to a file and reload it later:

```r
write_blueprint(bp, "path/to/save/bp.rds")
bp <- read_blueprint("path/to/save/bp.rds")
```

To find out more about creating blueprints, check out [this vignette](https://glycoverse.github.io/glysmith/articles/blueprint.html).
You can also have AI to create a blueprint for you. Check out [this vignette](https://glycoverse.github.io/glysmith/articles/ai.html).

**Happy forging!**
